ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2025-06-01 paper/Manuscript-Data-Package-1Round-Review.tex
Create a document with all manuscript changes and add answer letters for the reviewers.

\begin{abstract}
This paper describes \{ActiveCA\}, an open data product with Canadian
time use data. \{ActiveCA\} is an \texttt{R} data package that contains
analysis-ready data related to active travel spanning almost 40 years,
extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24
(2010), 29 (2015), and 34 (2022) of the Time Use Survey (TUS) cicles
from the General Social Survey (GSS). Active travel is characterized by
mode, with walking being part of every cycle and bicycling starting in
1992. The attributes of active trips are the types of locations of
origins and destinations, the duration of trips, and episode weights for
expanding the trips to population-wide estimates. Based on the year of
the survey, a variety of locations are coded. In earlier cycles, these
include home, work or school, and other's home, whereas in later cycles
these are augmented with locations such as grocery stores, restaurants,
outdoor destinations, and others. The geographical resolution includes
the province and whether the episode was in an urban or rural setting.
\end{abstract}

@@

\{ActiveCA\} is distributed as an \texttt{R} package with a number of
data objects and their documentation. \texttt{R} packages contain code,
data, and documentation in a standardized format that can be installed
by \texttt{R} users via a software repository, such as CRAN
(Comprehensive R Archive Network) or GitHub, which makes them an adroit
medium to distribute analysis-ready data.

@@

\subsection{\texorpdfstring{The \emph{Main
file}}{The Main file}}\label{the-main-file}

@@

The \emph{Main file} of the 2015 GSS surveys includes a total of 17,390
respondents, representing 29,766,399 individuals and 848 variables. For
discrete variables, Statistics Canada has assigned specific codes to the
possible values, with each code accompanied by a label. For instance, in
the case of the variable \texttt{SURVMNTH}, a value of \texttt{1} means
\texttt{January\ 2016}, 2 means \texttt{February\ 2016}, 3 corresponds
to \texttt{March\ 2016}, and so on.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `WGHT\_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-main-file-2015}\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the Main file of the 2015 GSS.}\\
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & 7 & 5 & 1 & 5\\
10001 & 8516.6140 & 7 & 5 & 1 & 1\\
10002 & 371.7520 & 1 & 4 & 2 & 1\\
10003 & 1019.3135 & 3 & 6 & 2 & 5\\
10004 & 1916.0708 & 9 & 2 & 1 & 6\\
\addlinespace
10005 & 1952.2015 & 4 & 1 & 1 & 6\\
10006 & 5761.5528 & 8 & 1 & 1 & 6\\
10007 & 466.0426 & 6 & 5 & 2 & 3\\
10008 & 2479.2991 & 2 & 2 & 2 & 1\\
10009 & 1436.1641 & 8 & 6 & 1 & 3\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

\subsection{\texorpdfstring{The \emph{Episode
file}}{The Episode file}}\label{the-episode-file}

@@

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an
example of the \emph{Main file} structure, Table
\ref{tab:ep-2015-unprocessed} presents the first seven episodes for the
record identification number \texttt{10041} and some variables from the
TUS PUMF 2015 \emph{Episode file} (Cycle 29). Each row in the table
corresponds to an episode associated with the specified record
identification (\texttt{PUMFID\ =\ 10041}), including the episode's
weight (\texttt{WGHT\_EPI}), episode number (\texttt{EPINO}), activity
code (\texttt{TUI\_01}), episode duration (\texttt{DURATION}), and
episode location (\texttt{LOCATION}).

@@

In the case illustrated in Table \ref{tab:ep-2015-unprocessed}, this
respondent began the diary description by sleeping at home
(\texttt{TUI\_01\ =\ 1} and \texttt{LOCATION\ =\ 300}) for 210 minutes,
followed by 40 minutes of personal hygiene (\texttt{TUI\_01\ =\ 2}). The
respondent then spent 15 minutes on personal care activities, such as
getting ready for school, supervising homework, reading, playing,
reprimanding, or providing educational or emotional support, as
indicated by \texttt{TUI\_01\ =\ 27}. Next, they recorded a travel
episode, walking for 15 minutes (\texttt{TUI\_01\ =\ 7} and
\texttt{LOCATION\ =\ 315}), where both the origin and destination were
their home. Such trips, where the journey starts and finishes at home,
can be classified as recreational or leisure trips. Next, the respondent
spent 3 hours searching for a job (\texttt{TUI\_01\ =\ 9}), took a
15-minute lunch break (\texttt{TUI\_01\ =\ 6}), and then cleaned the
house (\texttt{TUI\_01\ =\ 18}) for two hours. Table
\ref{tab:ep-2015-unprocessed} displays only six variables out of the 527
available. As shown, since the dataset does not label the variable
values, decoding them can be both time-consuming and challenging.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-epi-file-2015}\label{tab:ep-2015-unprocessed}Visualization of the seven first episodes of the record number `10041`.}\\
\toprule
PUMFID & EPINO & WGHT\_EPI & TUI\_01 & DURATION & LOCATION\\
\midrule
10041 & 1 & 1353.818 & 1 & 210 & 300\\
10041 & 2 & 1353.818 & 2 & 40 & 300\\
10041 & 3 & 1353.818 & 27 & 15 & 300\\
10041 & 4 & 1353.818 & 7 & 15 & 315\\
10041 & 5 & 1353.818 & 9 & 180 & 300\\
\addlinespace
10041 & 6 & 1353.818 & 6 & 15 & 300\\
10041 & 7 & 1353.818 & 18 & 120 & 300\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

As output, the \{ActiveCA\} package provides processed datasets of
walking and cycling episodes by year, a unified dataset that contains
all walking and cycling episodes across TUS cycles, and processed
\emph{Main files} (not unified, as socioeconomic and demographic
variables and their categories differ across TUS cycles).

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `WGHT\_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-processed-file-2015}\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the 2015 TUS Main File.}\\
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & July & 55 to 64 years & Male & Divorced\\
10001 & 8516.6140 & July & 55 to 64 years & Male & Married\\
10002 & 371.7520 & January & 45 to 54 years & Female & Married\\
10003 & 1019.3135 & March & 65 to 74 years & Female & Divorced\\
10004 & 1916.0708 & September & 25 to 34 years & Male & Single, never married\\
\addlinespace
10005 & 1952.2015 & April & 15 to 24 years & Male & Single, never married\\
10006 & 5761.5528 & August & 15 to 24 years & Male & Single, never married\\
10007 & 466.0426 & June & 55 to 64 years & Female & Widowed\\
10008 & 2479.2991 & February & 25 to 34 years & Female & Married\\
10009 & 1436.1641 & August & 65 to 74 years & Male & Widowed\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{TableNotes}
\begin{longtable}[t]{rrlrlll}
\caption{\label{tab:gss-processed-file-2015}\label{tab:ep-2015-processed}Visualization of the active travel episode for the record number `10041` of the 2015 GSS survey.}\\
\toprule
PUMFID & WGHT\_EPI & Activity & Duration & Origin & Destination & Mode\\
\midrule
10041 & 1353.818 & Transport to or from activity & 15 & Home & Home & Walking\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\begin{table}
\centering
\caption{\label{tab:cma-durations}\label{tab:cma-durations}Differences in walking duration between provinces and population centre type.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Population centre type} \\
\cmidrule(l{3pt}r{3pt}){2-3}
Province & CMA/CA & non CMA/CA\\
\midrule
Alberta & 15 & 5\\
British Columbia & 15 & 10\\
Manitoba & 10 & 10\\
New Brunswick & 5 & 5\\
Newfoundland and Labroador & 20 & 10\\
\addlinespace
Nova Scotia & 30 & 5\\
Ontario & 15 & 15\\
Prince Edward Island &  & 5\\
Quebec & 15 & 15\\
Saskatchewan & 10 & 5\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item 'CMA' denotes Census Metropolitan Area and 'CA' denotes Census agglomeration.
\end{tablenotes}
\end{threeparttable}}
\end{table}

@@

\{ActiveCA\} also provides a Jupyter Notebook containing a Python script
that demonstrates how to read R data files (.rda) and convert them into
Pandas DataFrames. This process allows users to work with and utilize
the datasets available in \{ActiveCA\} within a Python project.

@@

The value of \{ActiveCA\} lies in its transparency, accessibility, and
ease of use, which facilitates the addition of complementary data sets
in the future. \texttt{R} users can seamlessly explore TUS walking and
cycling episodes, with the option to suggest enhancements to the package
as needed. This article adopts the structure proposed by Anastasia and
Páez \citeyearpar{soukhov2023}, whose work provided essential guidance
for the creation of this package. Similarly, we aim to contribute to the
academic community by promoting transparent research practices that
encourage replication and innovation in related fields. We believe that
\{ActiveCA\} will serve as a basis for further research on TUS and for
the integration of additional data by the authors or the wider open
source community.

>>>>

\begin{abstract}
This paper describes \{ActiveCA\}, an open data product with Canadian
time use data. \{ActiveCA\} is an R data package that contains
analysis-ready data related to active travel spanning almost 40 years,
extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24
(2010), 29 (2015), and 34 (2022) of the Time Use Survey (TUS) cicles
from the General Social Survey (GSS). Active travel is characterized by
mode, with walking being part of every cycle and bicycling starting in
1992. The attributes of active trips are the types of locations of
origins and destinations, the duration of trips, and episode weights for
expanding the trips to population-wide estimates. Based on the year of
the survey, a variety of locations are coded. In earlier cycles, these
include home, work or school, and other's home, whereas in later cycles
these are augmented with locations such as grocery stores, restaurants,
outdoor destinations, and others. The geographical resolution includes
the province and whether the episode was in an urban or rural setting.
\end{abstract}

@@

\{ActiveCA\} is distributed as an \texttt{R} package with a number of
data objects and their documentation. \texttt{R} packages contain code,
data, and documentation in a standardized format that can be installed
by R users via a software repository, such as CRAN (Comprehensive R
Archive Network) or GitHub, which makes them an adroit medium to
distribute analysis-ready data.

@@

\subsection{The Main file}\label{the-main-file}

@@

The \emph{Main file} of the 2015 GSS surveys includes a total of 17,390
respondents, representing 29,766,399 individuals and 848 variables. For
discrete variables, Statistics Canada has assigned specific codes to the
possible values, with each code accompanied by a label. For instance, in
the case of the variable \texttt{SURVMNTH}, a value of \texttt{1} means
\texttt{January\ 2016}, \texttt{2} means \texttt{February\ 2016},
\texttt{3} corresponds to \texttt{March\ 2016}, and so on.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. WGHT\_PER:  person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-main-file-2015}\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the Main file of the 2015 GSS.}\\
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & 7 & 5 & 1 & 5\\
10001 & 8516.6140 & 7 & 5 & 1 & 1\\
10002 & 371.7520 & 1 & 4 & 2 & 1\\
10003 & 1019.3135 & 3 & 6 & 2 & 5\\
10004 & 1916.0708 & 9 & 2 & 1 & 6\\
\addlinespace
10005 & 1952.2015 & 4 & 1 & 1 & 6\\
10006 & 5761.5528 & 8 & 1 & 1 & 6\\
10007 & 466.0426 & 6 & 5 & 2 & 3\\
10008 & 2479.2991 & 2 & 2 & 2 & 1\\
10009 & 1436.1641 & 8 & 6 & 1 & 3\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

\subsection{The Episode file}\label{the-episode-file}

@@

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an
example of the \emph{Main file} structure, Table
\ref{tab:ep-2015-unprocessed} presents the first seven episodes for the
record identification number 10041 and some variables from the TUS PUMF
2015 \emph{Episode file} (Cycle 29). Each row in the table corresponds
to an episode associated with the specified record identification
(\texttt{PUMFID} = \texttt{10041}), including the episode's weight
(\texttt{WGHT\_EPI}), episode number (\texttt{EPINO}), activity code
(\texttt{TUI\_01}), episode duration (\texttt{DURATION}), and episode
location (\texttt{LOCATION}).

@@

In the case illustrated in Table \ref{tab:ep-2015-unprocessed}, this
respondent began the diary description by sleeping at home
(\texttt{TUI\_01\ =\ 1} and \texttt{LOCATION\ =\ 300}) for 210 minutes,
followed by 40 minutes of personal hygiene (\texttt{TUI\_01\ =\ 2}). The
respondent then spent 15 minutes on personal care activities, such as
getting ready for school, supervising homework, reading, playing,
reprimanding, or providing educational or emotional support, as
indicated by \texttt{TUI\_01\ =\ 27.} Next, they recorded a travel
episode, walking for 15 minutes (\texttt{TUI\_01\ =\ 7} and
\texttt{LOCATION\ =\ 315}), where both the origin and destination were
their home. Such trips, where the journey starts and finishes at home,
can be classified as recreational or leisure trips. Next, the respondent
spent 3 hours searching for a job (\texttt{TUI\_01\ =\ 9}), took a
15-minute lunch break (\texttt{TUI\_01\ =\ 6}), and then cleaned the
house (\texttt{TUI\_01\ =\ 18}) for two hours. Table
\ref{tab:ep-2015-unprocessed} displays only six variables out of the 527
available. As shown, since the dataset does not label the variable
values, decoding them can be both time-consuming and challenging.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. EPINO: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-epi-file-2015}\label{tab:ep-2015-unprocessed}Visualization of the seven first episodes of the record number 10041.}\\
\toprule
PUMFID & EPINO & WGHT\_EPI & TUI\_01 & DURATION & LOCATION\\
\midrule
10041 & 1 & 1353.818 & 1 & 210 & 300\\
10041 & 2 & 1353.818 & 2 & 40 & 300\\
10041 & 3 & 1353.818 & 27 & 15 & 300\\
10041 & 4 & 1353.818 & 7 & 15 & 315\\
10041 & 5 & 1353.818 & 9 & 180 & 300\\
\addlinespace
10041 & 6 & 1353.818 & 6 & 15 & 300\\
10041 & 7 & 1353.818 & 18 & 120 & 300\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

As output, the \{ActiveCA\} package provides processed datasets of
walking and cycling episodes by year, a unified dataset that contains
all walking and cycling episodes across TUS cycles, and processed main
files (not unified, as socioeconomic and demographic variables and their
categories differ across TUS cycles).

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. WGHT\_PER:  person weight. SURVMNTH: survey month of data collection. AGEGR10: age group of the respondent. SEX: sex of the respondent. MARSTAT: marital status of the respondent.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-processed-file-2015}\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the 2015 TUS Main File.}\\
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & July & 55 to 64 years & Male & Divorced\\
10001 & 8516.6140 & July & 55 to 64 years & Male & Married\\
10002 & 371.7520 & January & 45 to 54 years & Female & Married\\
10003 & 1019.3135 & March & 65 to 74 years & Female & Divorced\\
10004 & 1916.0708 & September & 25 to 34 years & Male & Single, never married\\
\addlinespace
10005 & 1952.2015 & April & 15 to 24 years & Male & Single, never married\\
10006 & 5761.5528 & August & 15 to 24 years & Male & Single, never married\\
10007 & 466.0426 & June & 55 to 64 years & Female & Widowed\\
10008 & 2479.2991 & February & 25 to 34 years & Female & Married\\
10009 & 1436.1641 & August & 65 to 74 years & Male & Widowed\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: PUMFID: record identification. EPINO: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{TableNotes}
\begin{longtable}[t]{rrlrlll}
\caption{\label{tab:gss-processed-file-2015}\label{tab:ep-2015-processed}Visualization of the active travel episode for the record number 10041 of the 2015 GSS survey.}\\
\toprule
PUMFID & WGHT\_EPI & Activity & Duration & Origin & Destination & Mode\\
\midrule
10041 & 1353.818 & Transport to or from activity & 15 & Home & Home & Walking\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\begin{table}
\centering
\caption{\label{tab:cma-durations}\label{tab:cma-durations}Differences in walking duration between provinces and population centre type.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Population centre type} \\
\cmidrule(l{3pt}r{3pt}){2-3}
Province & CMA/CA & non CMA/CA\\
\midrule
Alberta & 15 & 5\\
British Columbia & 15 & 10\\
Manitoba & 10 & 10\\
New Brunswick & 5 & 5\\
Newfoundland and Labroador & 20 & 10\\
\addlinespace
Nova Scotia & 30 & 5\\
Ontario & 15 & 15\\
Prince Edward Island &  & 5\\
Quebec & 15 & 15\\
Saskatchewan & 10 & 5\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item CMA denotes Census Metropolitan Area and CA denotes Census agglomeration.
\end{tablenotes}
\end{threeparttable}}
\end{table}

@@

\{ActiveCA\} also provides a Jupyter Notebook containing a
\texttt{Python} script that demonstrates how to read \texttt{R} data
files (.rda) and convert them into Pandas DataFrames. This process
allows users to work with and utilize the datasets available in
\{ActiveCA\} within a \texttt{Python} project.

@@

The value of \{ActiveCA\} lies in its transparency, accessibility, and
ease of use, which facilitates the addition of complementary data sets
in the future. R users can seamlessly explore TUS walking and cycling
episodes, with the option to suggest enhancements to the package as
needed. This article adopts the structure proposed by Anastasia and Páez
\citeyearpar{soukhov2023}, whose work provided essential guidance for
the creation of this package. Similarly, we aim to contribute to the
academic community by promoting transparent research practices that
encourage replication and innovation in related fields. We believe that
\{ActiveCA\} will serve as a basis for further research on TUS and for
the integration of additional data by the authors or the wider open
source community.
